---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading necessary packages

In this report, I will use `dplyr` package for modifying and analyzing the
data set, and the `ggplot2` package to display plot the data, and `lubridate`
package to deal with dates.

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
```

## Loading and preprocessing the data

```{r Loading Data}
df <- read.csv("activity.csv")
dim(df)
summary(df)
```

Have to convert the date column from character class to Date class

```{r}
df <- transform(df, date = ymd(date))
summary(df)
```

## What is mean total number of steps taken per day?

### Calculating total number of steps taken per day

```{r Mean Steps Per Day}
tbl_mean_steps <- df %>% group_by(date) %>%
                    summarise(total.steps = sum(steps, na.rm = TRUE))
```

### Histogram of total number of steps taken each day

```{r}
g <- ggplot(data = tbl_mean_steps)
g + geom_histogram(mapping = aes(x = total.steps), binwidth = 2500,
                   fill = "steelblue") +
    labs(title = "Histogram of Total Steps Taken Per Day",
         x = "Total Steps", y = "Count")
```

The mean and median of total number of steps taken per day are as follows:

```{r}
mn0 <- round(mean(tbl_mean_steps$total.steps), digits = 2)
md0 <- round(median(tbl_mean_steps$total.steps), digits = 2)

mn0
md0
```


## What is the average daily activity pattern?

### Making a time series plot

```{r}
ts_data <- df %>% group_by(interval) %>%
            summarise(mean.steps = mean(steps, na.rm = TRUE))
g <- ggplot(data = ts_data, mapping = aes(x = interval, y = mean.steps))
g + geom_line(colour = "purple") +
    labs(title = "Time Series Plot of Average Steps Taken vs Interval",
         x = "Interval (Minute)", y = "Average Steps")
```

The 5-minute interval starting at
`r subset(as.data.frame(ts_data), mean.steps == max(mean.steps))$interval` minute
contains the maximum number of steps, averaged across all days.

## Imputing missing values

```{r}
sum(is.na(df$steps))
```

We can see that there are total of `r sum(is.na(df$steps))` missing values in the
data. We will replace them by the average number of steps taken during that interval.

```{r}
# Separate the entries with missing values from the others
df_na <- df[is.na(df$steps), ]
df_not_na <- df[!is.na(df$steps), ]

# Create a merged data frame using "interval" column
mrg <- merge(df_na, ts_data, by = "interval") %>% mutate(steps = mean.steps) %>%
        select(steps, date, interval)

# Create the new data frame with zero missing values
df_new <- rbind(mrg, df_not_na) %>% arrange(date, interval)

# Verify that there are no missing values
sum(is.na(df_new$steps))
```

### Histogram of total number of steps taken each day

```{r Mean Steps Per Day New Data}
tbl_mean_steps_new <- df_new %>% group_by(date) %>%
                    summarise(total.steps = sum(steps, na.rm = TRUE))

# Drawing Plot
g <- ggplot(data = tbl_mean_steps_new)
g + geom_histogram(mapping = aes(x = total.steps), binwidth = 2500,
                   fill = "steelblue") +
    labs(title = "Histogram of Total Steps Taken Per Day",
         x = "Total Steps", y = "Count")
```

The mean and median of total number of steps taken per day are as follows:

```{r}
mn1 <- round(mean(tbl_mean_steps_new$total.steps), digits = 2)
md1 <- round(median(tbl_mean_steps_new$total.steps), digits = 2)

mn1
md1
```

Imputing the data has changed the mean and median from `r mn0` and `r md0` from
first section to `r mn1` and `r md1` respectively. Change in the total daily number
of steps taken can be observed as follows:
```{r}
total0 <- sum(tbl_mean_steps$total.steps)
total1 <- sum(tbl_mean_steps_new$total.steps)

total0
total1
```

## Are there differences in activity patterns between weekdays and weekends?

```{r}
df_new <- df_new %>% mutate(day.of.week = wday(date, label = TRUE))
dim(df_new)
head(df_new)

# Create a new vector indicating whether the day.of.month is a weekday or a
# weekend
day <- mapply(df_new$day.of.week,
              FUN = function(x){if(x %in% c("Sat","Sun")){"weekend"}
                  else{"weekday"}})

df_new <- df_new %>% mutate(day = factor(day))

# Calculate average number of steps taken across intervals
df_new <- df_new %>% group_by(interval, day) %>% mutate(mean.steps = mean(steps))

df_new

# Plot tie series
g <-  ggplot(data = df_new, mapping = aes(x = interval, y = mean.steps))
g + geom_line(colour = "blue") + facet_grid(day ~ .) +
    labs(title = "Time Series Plot of Average Steps Taken vs Interval",
         x = "Interval", y = "Average Steps")
```
